Supporting information for ``Environmental drivers of synchrony and their interactions''

### 2022-11-08 Paper/SupMat.Rnw
various edits to Discussion and getting repo ready
changes to the writeup from a long time ago that I forgot to commit. comming back to this for the first time in a while now.
mainly added a bunch of comments which suggest how to restructure the write-up for simplicity, which is what I am going to do next

%header material for paper
\title{Supporting information for ``How environmental drivers of synchrony interact''}
\date{}

@@

The following examples should help clarify the properties of our definition of cross synchrony.
For pedagogical clarity, we start with a white-noise example that does not illustrate the full value of the
idea of cross synchrony, but which is simple; we subsequently move on to an example which is more
complex but also more illustrative. These examples are also used in the theoretical case studies (main text).

@@

We now explore why the results above for $\rho_{\epsilon^{(1)}\epsilon^{(1)}}$, $\rho_{\epsilon^{(2)}\epsilon^{(2)}}$ and
$\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ make sense for the noise processes we described, particularly for the sets of parameters 
used for theoretical case study C in the main text (see Fig. \ref{fig:rhosForC}). 
The pre-factor $\left| \frac{1}{1-c_1 \mu - c_2 \mu^2} \right|^2$ in all these expressions reflects the autoregressive 
manner in which the noises $\epsilon$ were defined. Autoregressive models of order $2$ are well known to be capable of
producing oscillations for certain values of the parameters. For instance, for case study C, we 
use $c_1=\Sexpr{ns_C1[1]}$ and $c_2=\Sexpr{round(ns_C1[2],4)}$, values
which cause the expression $\left| \frac{1}{1-c_1 \mu - c_2 \mu^2} \right|^2$ to have a single peak at
frequency $0.25$ cycles per sampling interval. It makes sense that the synchrony quantities
$\rho_{\epsilon^{(1)}\epsilon^{(1)}}$, $\rho_{\epsilon^{(2)}\epsilon^{(2)}}$ and
$\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ for periodically oscillating quantities should also have a peak at the 
timescale of oscillation. 
The term $\exp(-2 \pi \iota f (l_1-l_2))$ in the expression for 
$\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ reflects that fact that cross variable synchrony for this example is lagged
by an amount $l_1-l_2$. The lack of a similar term in the expressions for $\rho_{\epsilon^{(1)}\epsilon^{(1)}}$
or $\rho_{\epsilon^{(2)}\epsilon^{(2)}}$ reflects that fact that there are no lags built into within-variable
synchrony in this example.

@@

Floating surface canopy of kelp was measured by remote sensing, with details and
validation reported by \cite{Cavanaugh2011} and \cite{Bell2020, Bell2021}. Multispectral Landsat
imagery of $30$m resolution was used. Quarterly mean biomass values and spatial extent of canopy  
within $500$m coastline segments (henceforth called ``locations'')
were computed by averaging. Quarters were: Q1, January-March; Q2, April-June; 
Q3, July-September; Q4, October-December. 
Data were thusly available from $\Sexpr{dim(kelp_SB)[1]}$ locations spanning 
approximately from Monterey Bay to San Diego. 
We then chose locations for analysis for which kelp was ``persistent'', in this 
context meaning absent or missing data for no more than three years; this was to focus on 
population dynamics of kelp rather than extinction-recolonization dynamics, which occur in some 
locations and which can relate to sediment dynamics on the ocean floor and other factors
not currently of primary interest. Of the original $\Sexpr{dim(kelp_SB)[1]}$ locations,
$\Sexpr{dim(kelp_MI)[1]}$ ($\Sexpr{100*round(dim(kelp_MI)[1]/dim(kelp_SB)[1],2)}$\%) 
were deemed persistent according to this criterion. Missing values in the remaining 
time series were then filled with seasonal medians. The biomass and spatial extent time series
for each location were then used together to produce one time series estimating biomass
density per unit useable habitat at the site, as follows: every value of 
the biomass time series was divided by the $90^{\text{th}}$ quantile of the distribution
of the spatial extent time series at the site. The $90^{\text{th}}$ quantile was used instead
of the maximum value of the spatial extent time series because maximum statistics have high variance. 
This procedure served to normalize total biomass by an estimate
of total useable habitat at a site. Kelp establish on rocky substrate, and areas with sandy or
sedimented substrate are not useable kelp habitat. All time series were then linearly detrended,
and only those time series were utilized which were located within the three regions pictured
in Fig. \ref{MT-fig:map} in the main text. Regions were selected to represent distinct dynamical
patterns known to occur in central and southern CA, and also to have another the same number
of time series per selected region.

Wave nutrient data were also available for each quarter at each location. Wave data were derived 
by applying a swell propagation model with bouy-based wave measurements and a validation 
techniques \citep{Wingeart2001,Hansen2009,Bell2015,OReilly2016}. Data on nitrates 
were produced \citep{Snyder2020} by exploiting an established empirical 
relationship between surface nitrate concentration
and sea surface temperature (SST) in coastal CA \citep{Zimmerman1984,Palacios2013,Snyder2020}
and SST data derived from Advanced Very High Resolution Radiometer satellite imagery. 
Wave and nutrient time series in each location were again linearly detrended before use.

@@

\section{Supporting results for kelp}

>>>>

%header material for paper
\title{Supporting information for ``Environmental drivers of synchrony and their interactions''}
\date{}

@@

The following examples should help clarify the properties of our definition of cross synchrony.
For pedagogical clarity, we start with a white-noise example that does not illustrate the full value of the
idea of cross synchrony, but which is simple; we then move on to an example which is more
complex but also more illustrative. These examples are also used in the theoretical case studies (main text).

@@

We now explore why the results above for $\rho_{\epsilon^{(1)}\epsilon^{(1)}}$, $\rho_{\epsilon^{(2)}\epsilon^{(2)}}$ and
$\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ make sense for the noise processes we described, particularly for the sets of parameters 
used for theoretical case study C in the main text (see Fig. \ref{fig:rhosForC}). 
The pre-factor $\left| \frac{1}{1-c_1 \mu - c_2 \mu^2} \right|^2$ in all these expressions reflects the autoregressive 
manner in which the noises $\epsilon$ were defined. Autoregressive models of order $2$ are well known to be capable of
producing oscillations for certain values of the parameters. For instance, for case study C, we 
use $c_1=\Sexpr{ns_C1[1]}$ and $c_2=\Sexpr{round(ns_C1[2],4)}$, values
which cause the expression $\left| \frac{1}{1-c_1 \mu - c_2 \mu^2} \right|^2$ to have a single peak at
frequency $0.25$ cycles per sampling interval. It makes sense that the synchrony quantities
$\rho_{\epsilon^{(1)}\epsilon^{(1)}}$, $\rho_{\epsilon^{(2)}\epsilon^{(2)}}$ and
$\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ for periodically oscillating quantities should also have a peak at the 
timescale of oscillation. 
The term $\exp(-2 \pi \iota f (l_1-l_2))$ in the expression for 
$\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ reflects the fact that cross variable synchrony for this example is lagged
by an amount $l_1-l_2$. The lack of a similar term in the expressions for $\rho_{\epsilon^{(1)}\epsilon^{(1)}}$
or $\rho_{\epsilon^{(2)}\epsilon^{(2)}}$ reflects that fact that there are no lags built into within-variable
synchrony in this example.

@@

Floating surface canopy of kelp was measured by remote sensing, with details and
validation reported by \cite{Cavanaugh2011} and \cite{Bell2020, Bell2021}. Multispectral Landsat
imagery of $30$m resolution was used. Quarterly mean biomass values and spatial extent of canopy  
within $500$m coastline segments (henceforth called ``locations'')
were computed by averaging. Quarters were: Q1, January-March; Q2, April-June; 
Q3, July-September; Q4, October-December. 
Data were thusly available from $\Sexpr{dim(kelp_SB)[1]}$ locations spanning 
approximately from Monterey Bay to San Diego. 
We then chose locations for analysis for which kelp was ``persistent'', in this 
context meaning absent or missing data for no more than three years; this was to focus on 
population dynamics of kelp rather than extinction-recolonization dynamics, which occur in some 
locations and which can relate to sediment dynamics on the ocean floor and other factors
not currently of primary interest. Of the original $\Sexpr{dim(kelp_SB)[1]}$ locations,
$\Sexpr{dim(kelp_MI)[1]}$ ($\Sexpr{100*round(dim(kelp_MI)[1]/dim(kelp_SB)[1],2)}$\%) 
were deemed persistent according to this criterion. Missing values in the remaining 
time series were then filled with seasonal medians. The biomass and spatial extent time series
for each location were then used together to produce one time series estimating biomass
density per unit useable habitat at the site, as follows: every value of 
the biomass time series was divided by the $90^{\text{th}}$ quantile of the distribution
of the spatial extent time series at the site. The $90^{\text{th}}$ quantile was used instead
of the maximum value of the spatial extent time series because maximum statistics have high variance. 
This procedure served to normalize total biomass by an estimate
of total useable habitat at a site. Kelp establish on rocky substrate, and areas with sandy or
sedimented substrate are not useable kelp habitat. All time series were then linearly detrended,
and only those time series were utilized which were located within the three regions pictured
in Fig. \ref{MT-fig:map} in the main text. Regions were selected to represent distinct dynamical
patterns known to occur in central and southern CA, and also to have about the same number
of time series per selected region.

Wave and nutrient data were also available for each quarter at each location. Wave data were derived 
by applying a swell propagation model with bouy-based wave measurements and a validation 
techniques \citep{Wingeart2001,Hansen2009,Bell2015,OReilly2016}. Data on nitrates 
were produced \citep{Snyder2020} by exploiting an established empirical 
relationship between surface nitrate concentration
and sea surface temperature (SST) in coastal CA \citep{Zimmerman1984,Palacios2013,Snyder2020}
and SST data derived from Advanced Very High Resolution Radiometer satellite imagery. 
Wave and nutrient time series in each location were again linearly detrended before use.

@@

\clearpage
\newpage
